---
title: "DATA 607 - TidyVerse Vignette"
author: "Catherine Dube"
output: 
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    code_folding: hide
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, scales)
```

## Set Up
I am using the data available at [the U.S. Bureau of Labor Statistics](https://www.bls.gov/cps/cps_aa2023.htm) to look at the 2023 earnings of full-time wage and salary workers in the U.S. I've hosted the table of data I am interested in in my personal repo.

## Read in Data
```{r read-in-data}
df_raw <- read.csv('https://raw.githubusercontent.com/cdube89128/DATA-607/refs/heads/main/week-09/cpsa2023_table39.csv')
head(df_raw)
sapply(df_raw, class)
```

This data is barely readable as is, so let's take it to some substantial cleaning.

## Tidy the Data
```{r org-data}
new_colnames <- c("metric", "total_number_workers", "median_weekly_earnings", 
                  "men_number_workers", "men_weekly_earnings",
                  "women_number_workers", "women_weekly_earnings")
df <- df_raw[7:nrow(df_raw),]
colnames(df) <- new_colnames

# Remove empty rows
df <- df %>%
  filter(if_all(everything(), ~ . != "" & !is.na(.)))

# Clean up the dashes, dollar signs, and make numeric instead of char
df <- df %>%
  mutate(across(
    -metric,
    ~ str_replace_all(., "[$,]", "") |>   # remove $ and commas
      na_if("–") |>                       # convert en dash to NA
      as.numeric()                        # convert to numeric
  ))

summary(df)
```

## Taking A Look at Men's and Women's Earnings
```{r men-women-earnings-explore}
df_gap <- df %>%
  filter(!is.na(men_weekly_earnings), !is.na(women_weekly_earnings)) %>%  # keep only rows with data for both men +women
  mutate(
    pay_gap = men_weekly_earnings - women_weekly_earnings,
    women_to_men_ratio = women_weekly_earnings / men_weekly_earnings,
    women_percent_of_men = women_to_men_ratio * 100
  ) %>%
  arrange(women_to_men_ratio)
```

## Jobs Where Women Earn the Smallest Percentage of What Men Earn
```{r big-gaps}
df_gap %>%
  select(metric, women_percent_of_men) %>%
  arrange(women_percent_of_men) %>%
  head(10)
```

## Visualizations
```{r visual}
df_gap_filtered <- df_gap %>%
  filter(!is.na(women_to_men_ratio)) %>%
  arrange(women_to_men_ratio)

df_gap_topbottom <- bind_rows(
  head(df_gap_filtered, 10),
  tail(df_gap_filtered, 10)
)

ggplot(df_gap_topbottom, aes(x = reorder(metric, women_to_men_ratio), y = women_to_men_ratio)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Occupations with the Largest and Smallest Gender Pay Gaps",
    x = "Occupation",
    y = "Women's Earnings as % of Men's"
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(hjust = 1)) 

```


## Overall Gender Pay Gap
```{r Gender Pay Gap}
df_overall <- df_gap %>%
mutate(
total_workers = men_number_workers + women_number_workers
) %>%
summarise(
total_men_workers = sum(men_number_workers, na.rm = TRUE),
total_women_workers = sum(women_number_workers, na.rm = TRUE),
avg_men_earnings_weighted =
sum(men_weekly_earnings * men_number_workers, na.rm = TRUE) /
total_men_workers,
avg_women_earnings_weighted =
sum(women_weekly_earnings * women_number_workers, na.rm = TRUE) /
total_women_workers,
women_to_men_ratio_overall =
avg_women_earnings_weighted / avg_men_earnings_weighted,
women_percent_of_men_overall = women_to_men_ratio_overall * 100
)

df_overall

```