---
title: "Corporate Profitability Analysis Using the Tidyverse"
author: "Escarlet Gabriel Vicente"
date: "2025-10-26"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Introduction

This vignette demonstrates how to use the Tidyverse to analyze corporate financial data from the Fortune Top 1000 Companies by Revenue (2022) dataset, obtained from Kaggle.
The goal is to explore business metrics such as revenue, profit, and profit margin, and identify which companies generate the highest profits relative to their revenues.

**Data Source:** [Fortune Top 1000 Companies by Revenue (2022), Kaggle](https://www.kaggle.com/datasets/surajjha101/fortune-top-1000-companies-by-revenue-2022)  
**License:** CC BY-NC-SA 4.0


---

## Load Libraries
We start by loading the Tidyverse package, which contains tools for data manipulation (dplyr), data cleaning (tidyr), and visualization (ggplot2).
```{r}
library(tidyverse)
```

## Import Dataset
The dataset is read into R using read_csv(). We then preview the first few rows to understand the structure of the data.
```{r}
fortune <- read_csv("Fortune 1000 Companies by Revenue.csv")

head(fortune)
```

## Clean and Prepare Data
The dataset includes numeric values stored as text, such as revenues and profits with $ and commas.
These symbols are removed, and the columns are converted into numeric format. Missing values are filtered out to ensure clean analysis.
```{r}
fortune <- fortune %>%
mutate(
revenues = as.numeric(gsub("[$,]", "", revenues)),
profits = as.numeric(gsub("[$,]", "", profits)),
employees = as.numeric(gsub(",", "", employees))
) %>%
filter(!is.na(revenues), !is.na(profits))

head(fortune)
```

## Calculate Profit Margin
We create a new variable, *profit_margin*, to measure the percentage of profit earned for every dollar of revenue.
This metric allows us to assess how efficiently companies convert sales into profit.
```{r}
fortune_clean <- fortune %>%
filter(revenues > 0) %>%
mutate(profit_margin = round((profits / revenues) * 100, 2))

head(fortune_clean)
```

## Top 10 Companies by Revenue
To identify the largest corporations in the dataset, we select the top 10 companies ranked by total revenue.
The table displays each company’s rank, name, total revenue, profit, and calculated profit margin.
```{r}
top10_revenue <- fortune_clean %>%
slice_max(order_by = revenues, n = 10) %>%
select(rank, name, revenues, profits, profit_margin)

top10_revenue
```

## Visualization 1: Revenue vs. Profit
This scatter plot compares revenue and profit among the top 30 companies.
It shows how high revenues do not always guarantee high profits, revealing differences in efficiency between companies.
```{r}
fortune_clean %>%
slice_max(order_by = revenues, n = 30) %>%
ggplot(aes(x = revenues, y = profits, color = name)) +
geom_point(size = 3, alpha = 0.8) +
labs(
title = "Revenue vs Profit for Top 30 Companies (Fortune 1000 - 2022)",
x = "Revenue (USD Millions)",
y = "Profit (USD Millions)",
color = "Company"
) +
theme_minimal()
```

## Visualization 2: Top Companies by Profit Margin
This bar chart highlights the top 10 companies with the highest profit margins, showing which firms are most successful at converting revenue into profit.
It provides insight into which companies achieve efficiency over scale.
```{r}
top10_margin <- fortune_clean %>%
slice_max(order_by = profit_margin, n = 10)

ggplot(top10_margin, aes(x = reorder(name, profit_margin), y = profit_margin, fill = name)) +
geom_col(show.legend = FALSE) +
coord_flip() +
labs(
title = "Top 10 Companies by Profit Margin (Fortune 1000 - 2022)",
x = "Company",
y = "Profit Margin (%)"
) +
theme_minimal()
```

## Insights and Conclusion
The results indicate a clear distinction between firms leading in revenue and those excelling in profitability. While Walmart and Amazon remain market leaders by scale, firms like Bio-Rad Laboratories and eBay stand out for their efficient profit generation. Through this analysis, the Tidyverse proved effective for structuring, transforming, and visualizing corporate financial data.