---
title: "TidyVerse Create"
author: "James Chun"
date: "October 25, 2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Here I will be working with two Kaggle datasets sourced from https://www.kaggle.com/datasets/sidtwr/videogames-sales-dataset/data 
<br> 
<br>
The two datasets contains information for the highest video game sales from the 2010's for the PS4 and Xbox One. The data also contains the publisher, genre, date released, and the sales for various counts (in units per million). 


## Load Libraries
```{r message=FALSE}
library(tidyverse)
library(ggplot2)
library(RCurl)
```


## Import Data
Data will be stored in an external github and accessed virtually. Note again that the units for the sales per country are in millions of copies sold.

```{r}
PS4_sales <- getURL("https://media.githubusercontent.com/media/Ryungje/DATA607/refs/heads/main/TidyVerse%20Create%20Project/PS4_GamesSales.csv") %>%
  read.csv(text = .)

XboxOne_sales <- getURL("https://media.githubusercontent.com/media/Ryungje/DATA607/refs/heads/main/TidyVerse%20Create%20Project/XboxOne_GameSales.csv") %>%
  read.csv(text = .)
  
```

## Data Tidying
For the most part, the datasets are "tidy" and not much work is needed. However, I will edit the sales columns to be in millions rather than a unit rate. This will be done using the mutate() function.

```{r}
PS4_sales <- PS4_sales %>%
  mutate(North.America = North.America * 1000000, 
         Europe = Europe * 1000000,
         Japan = Japan * 1000000,
         Rest.of.World = Rest.of.World * 1000000,
         Global = Global * 1000000)


XboxOne_sales <- XboxOne_sales %>%
  mutate(North.America = North.America * 1000000, 
         Europe = Europe * 1000000,
         Japan = Japan * 1000000,
         Rest.of.World = Rest.of.World * 1000000,
         Global = Global * 1000000)
```

## Exploratory Data Analysis
First, sort the datasets for highest sales globally and observe the head.

```{r}
XboxOne_sales %>%
  arrange(desc(Global)) %>% 
  head()
```
From this we can see that for the Xbox One, GTA V holds the highest global sales, followed by CoD Black Ops 3, CoD WWII and Red Dead Redemption 2. 

```{r}
PS4_sales %>%
  arrange(desc(Global)) %>% 
  head()
```
Meanwhile, for the PS4, the same is mostly true except Red Dead Redemption 2 takes the third spot from CoD WWII.

### Graphs 
Simple visuals created using graphing functions from ggplot2.

```{r}
XboxOne_sales %>%
  # Count top 5 sales by publisher
  group_by(Publisher) %>%
  summarise(total_sales = sum(Global, na.rm = TRUE)) %>%
  arrange(desc(total_sales)) %>%
  head(5) %>%
  
  # Make plot
  ggplot(aes(x = reorder(Publisher, total_sales), y = total_sales)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  
  # Customize plot
  labs(
    title = "Top Publishers by Global Sales for Xbox One",
    x = "Publisher",
    y = "Total Sales"
  ) +
  scale_y_continuous(labels = function(x) paste0(x / 1e6, "M")) +
  theme_minimal()
```


```{r}
PS4_sales %>%
  # Count top 5 sales by publisher
  group_by(Publisher) %>%
  summarise(total_sales = sum(Global, na.rm = TRUE)) %>%
  arrange(desc(total_sales)) %>%
  head(5) %>%
  
  # Make plot
  ggplot(aes(x = reorder(Publisher, total_sales), y = total_sales)) +
  geom_bar(stat = "identity", fill = "tomato") +
  coord_flip() +
  
  # Customize plot
  labs(
    title = "Top Publishers by Global Sales for PS4",
    x = "Publisher",
    y = "Total Sales"
  ) +
  scale_y_continuous(labels = function(x) paste0(x / 1e6, "M")) +
  theme_minimal()
```
Again, the data for the two consoles appear to be similar. However, Sony Interactive makes a notable contribution to global sales for the PS4 console. This is interesting and could be the result of Sony exclusives on the PS4: another motivation for people to buy the PS4.

## Extension: Genre-Based Sales Comparison (By Arutam Antunish)
We’ll compare total global sales by genre for both consoles. This adds a new dimension to the analysis and uses group_by(), summarise(), mutate(), and ggplot().

### Global Sales by Genre – Xbox One

```{r}
XboxOne_sales %>%
  group_by(Genre) %>%
  summarise(total_sales = sum(Global, na.rm = TRUE)) %>%
  arrange(desc(total_sales)) %>%
  ggplot(aes(x = reorder(Genre, total_sales), y = total_sales)) +
  geom_bar(stat = "identity", fill = "darkcyan") +
  coord_flip() +
  labs(title = "Global Sales by Genre – Xbox One",
       x = "Genre",
       y = "Total Sales") +
  scale_y_continuous(labels = function(x) paste0(x / 1e6, "M")) +
  theme_minimal()
```

### Global Sales by Genre – PS4

```{r}
PS4_sales %>%
  group_by(Genre) %>%
  summarise(total_sales = sum(Global, na.rm = TRUE)) %>%
  arrange(desc(total_sales)) %>%
  ggplot(aes(x = reorder(Genre, total_sales), y = total_sales)) +
  geom_bar(stat = "identity", fill = "darkorange") +
  coord_flip() +
  labs(title = "Global Sales by Genre – PS4",
       x = "Genre",
       y = "Total Sales") +
  scale_y_continuous(labels = function(x) paste0(x / 1e6, "M")) +
  theme_minimal()
```

These charts reveal which genres dominate global sales for each console. For example, if Action or Shooter games lead on both platforms, it may reflect broader player preferences across the industry.

## Summary

###EDIT
#Extended the analysis by comparing publishers across both consoles.
First combined the PS4 and Xbox One datasets into a single table, then look at the
publishers with the highest global sales. Finally created visualizations that
compare publisher performance by console.
# Combine PS4 and Xbox One sales data into a single table

combined_sales <- bind_rows(
PS4_sales %>%
select(Title, Publisher, Genre, North.America, Europe, Japan, Rest.of.World, Global) %>%
  mutate(Console = "PS4"),
  XboxOne_sales %>%
# Same structure for Xbox One
select(Title, Publisher, Genre, North.America, Europe, Japan, Rest.of.World, Global) %>%
  mutate(Console = "Xbox One")
)

# Combined dataset

head(combined_sales)

#Calculate total global sales by Publisher and Console

publisher_global <- combined_sales %>%
group_by(Publisher, Console) %>%
summarise(total_global = sum(Global, na.rm = TRUE), .groups = "drop")

#Find the top publishers overall (across both consoles)

top_publishers_overall <- publisher_global %>%
group_by(Publisher) %>%


summarise(overall_global = sum(total_global), .groups = "drop") %>%
arrange(desc(overall_global)) %>%

slice_head(n = 10)   # Top 10

#Keep top publishers in the detailed table

publisher_top <- publisher_global %>%
inner_join(top_publishers_overall, by = "Publisher")

# View the top publishers table

publisher_top

# Plot: For each of the top publishers, compare PS4 vs Xbox One

ggplot(publisher_top, aes(x = reorder(Publisher, overall_global),
  y = total_global,
  fill = Console)) +
  geom_col(position = "dodge") +  # Side-by-side bars
  coord_flip() +                  # Flip for better readability
  labs(title = "Top Publishers by Global Sales: PS4 vs Xbox One",
    x = "Publisher",
    y = "Total Global Sales (Units)",
    fill = "Console"
) +
###EDIT

In this vignette, we explored some basic data tidying and analysis using the libraries and functions:
<br>
- ***tidyverse*** for data tidying (summarise, group_by, arrange, mutate).  
- ***RCurl*** to import data (getURL).  
- ***ggplot2*** to visualize data (ggplot).  

